---
title: "Duncan Ray - Native Scots Pine and Climate Variables - 007c20 - 100401-1006"
author: "Toni Clarke"
date: "16 March 2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message=F, warning =F)
```

```{r}

library(tidyverse)
library(ggplot2)
library(lme4)
library(knitr)
library(lmerTest)
library(pdp)
#library("sjPlot", lib.loc="C:/R-Packages")

#setwd("/Users/toni.clarke/OneDrive - Forest Research/Documents/007c20/")
```

# Methods

\newline
Analysis was conducted in R (version 3.5.2, R Core Team 2018), with graphics produced using ggplot2 in R (Wickham, 2016). 

177 families of half sibs derived from 21 populations are grown on 3 different planting sites. Several climate variables are available and these are mapped to different planting sites - each planting site is assigned one value for a climate variable and so there are only ever 3 values for each variable. 

A linear model was initally built on a subset of pre-selected populations representing the most Northern/Southern sites etc (total of 33 families). The CC9 family was identified as having a significant interaction with AHM. 

This model was then applied to all families - excluding families where there were zero trees available at a planting site, and excluding all trees with a height of 0. This left 159 families and 1697 observations. The following model was fit and also applied to the PPT_wt variable. 

Height ~ Family + AHM + Family*AHM + (1|Population)


Differences between populations and the interaction with climate variables are also of interest. The following mixed linear model was fit, with population now fit as a fixed effect, and family as a random effect to account for similarities between half-sibs. AHM and PPT_wt were both analysed. 

Height ~ Population + AHM + Population*AHM + (1|Family)


```{r}

# Read in and clean data

  data <- read.csv("file:///C:/Users/vanessa.burton/OneDrive - Forest Research/Documents/R/DeltaTraitSDM/Scots_pine/Scotspine_Common_Garden_Heights_3.csv", header=T)

# Seqno and tag are unique identifiers
# Three planting sites - BORDERS (3), GLENSAUGH (1), INVEREWE (2)
# 4 Blocks, need to correct c to C
# 168 per block at each site but Inverewe does not have a block D 
  data$Block[data$Block=="c"]<-"C"
  data$Block <- factor(data$Block)
  
# 7 seed zones EC, NC NE, NN, NW, SC, SW
# ~ 96 of each seed zone at each site, but ~72 at Inverewe due to no Block D
# ~ 24 of each seed zone at each site in each block (24*4)
  
# 21 Different populations - 32 at each site (8 on each block)
# Each population is unique to a seed.zone, except GD which has 87 from EC and one from NE - error?
  data$Seed.Zone[data$Population=="GD"] <- "EC"
  
# Each seed zone has 3 populations derived from it.
  
# 177 different families, made up of provenance (population) name then the mother tree - between 4 and 11 entries for each family - these contain the half sibs

# Interested in how differen climate variables impact phenotype depending on family
# Wants mean line plots per family.
# Statistical significance could be determined using
# Pheno ~ family + climate + family*climate + seed.Zone + (1|population) + (1|blocksite)
# Using moisture index as example variable - AHM
  
# Make a list of families which have zero entries at any of the sites
# can filter analysis on tolose variable
  table_data <- data.frame(table(data$Family, data$PlantingSite))
  tolose <- table_data[table_data$Freq==0,1]
```



```{r}
# Make graphs with mean height per family plotted and then trend lines for AHM
# using loess smooth - Local Polynomial Regression Fitting
# Filter by population and facet_wrap by family


  F1 <- data %>%
  filter(Population =="CR") %>%
    ggplot(aes(x=AHM, y=W17Height, color=Family))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(16,30),breaks=c(17.5,23.44,27.54), labels=c("17.5", "23.4", "27.5")) +
    facet_wrap(~Family) +
    theme(legend.position="none")

  F2 <- data %>%
  filter(Population =="CR") %>%
    ggplot(aes(x=PPT_wt, y=W17Height, color=Family))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(175,350),breaks=c(182.4,218.9,342), labels=c("182.4", "218.9", "342")) +
    facet_wrap(~Family) +
    theme(legend.position="none")
  
    F5a <- data %>%
  filter(Family =="CR4") %>%
    ggplot(aes(x=PPT_wt, y=W17Height, color=Family))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(175,350),breaks=c(182.4,218.9,342), labels=c("182.4", "218.9", "342"))
    
    F5b <- data %>%
  filter(Family =="CR4") %>%
    ggplot(aes(x=AHM, y=W17Height, color=Family))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(16,30),breaks=c(17.5,23.44,27.54), labels=c("17.5", "23.4", "27.5"))

  F6 <- data %>%
    ggplot(aes(x=PPT_wt, y=W17Height, color=Population))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(175,350),breaks=c(182.4,218.9,342), labels=c("182.4", "218.9", "342")) +
    facet_wrap(~Population) +
    theme(legend.position="none")
  
    F7 <- data %>%
    ggplot(aes(x=AHM, y=W17Height, color=Population))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(16,30),breaks=c(17.5,23.44,27.54), labels=c("17.5", "23.4", "27.5")) +
    facet_wrap(~Population) +
    theme(legend.position="none")
    
    F8a <- data %>%
    filter(Population=="BE") %>%
    ggplot(aes(x=PPT_wt, y=W17Height, color=Population))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(175,350),breaks=c(182.4,218.9,342), labels=c("182.4", "218.9", "342")) +
    facet_wrap(~Population) +
    theme(legend.position="none")
    
    F8b <- data %>%
    filter(Population=="BE") %>%
    ggplot(aes(x=AHM, y=W17Height, color=Population))+
    geom_point()+
    stat_smooth() +
    ylim(0,30) +
    scale_x_continuous(limits=c(16,30),breaks=c(17.5,23.44,27.54), labels=c("17.5", "23.4", "27.5")) +
    facet_wrap(~Population) +
    theme(legend.position="none")
    

    

# Plot predicted regression lines by finding best fit model for AHM
# Only take data where there are trees present at every site (or not enough variation for linear model)
# Only take trees with height greater than zero

  clean_data <- data[!data$Family %in% tolose,]
  clean_data <- clean_data[clean_data$W17Height > 0, ]
  clean_data$Family <- factor(clean_data$Family)
  
# Model testing effect of Family and AHM and interaction
# 159 families, and 159 interaction

  m3 <- lmer(W17Height ~ Family + AHM + Family*AHM + (1|Population), data=clean_data)
  results <- data.frame(coef(summary(m3)))
  
  m3a <- lmer(W17Height ~ Family + PPT_wt + Family*PPT_wt + (1|Population), data=clean_data)
  results_a <- data.frame(coef(summary(m3a)))
  
# Take data with p-value less than 0.05
  results_sig <- results[results$`Pr...t..` <= 0.05,]
  
# Model testing effect of Population and AHM and interaction
  m4 <- lmer(W17Height ~ Population + AHM + Population*AHM + (1|Family), data=clean_data)
  results_m4 <- data.frame(coef(summary(m4)))
  
  m5 <- lmer(W17Height ~ Population + PPT_wt + Population*PPT_wt + (1|Family), data=clean_data)
  results_m5 <- data.frame(coef(summary(m5)))
  
  results_sigm4 <- results_m4[results_m4$`Pr...t..` <= 0.05,]
  results_sigm5 <- results_m5[results_m5$`Pr...t..` <= 0.05,]


  newdat1 <- expand.grid(
    PPT_wt=c(182.4,218.9,342)
    , Family=levels(clean_data$Family),
     W17Height = 0
)
  newdat1$W17Height <- predict(m3a,newdat1,re.form=NA)
  mm <- model.matrix(terms(m3a),newdat1)
  pvar1 <- diag(mm %*% tcrossprod(vcov(m3a),mm))
  tvar1 <- pvar1+VarCorr(m3a)$Population[1] # As population is ranef
  
  cmult <- 1.96 
newdat1 <- data.frame(
    newdat1
    , plo = newdat1$W17Height-cmult*sqrt(pvar1)
    , phi = newdat1$W17Height+cmult*sqrt(pvar1)
    , tlo = newdat1$W17Height-cmult*sqrt(tvar1)
    , thi = newdat1$W17Height+cmult*sqrt(tvar1)
)

  newdat <- expand.grid(
    AHM=c(17.5,23.44,27.54)
    , Family=levels(clean_data$Family),
     W17Height = 0
)
  newdat$W17Height <- predict(m3,newdat,re.form=NA)
  mm <- model.matrix(terms(m3),newdat)
  pvar1 <- diag(mm %*% tcrossprod(vcov(m3),mm))
  tvar1 <- pvar1+VarCorr(m3)$Population[1] # As population is ranef
  
  cmult <- 1.96 
newdat <- data.frame(
    newdat
    , plo = newdat$W17Height-cmult*sqrt(pvar1)
    , phi = newdat$W17Height+cmult*sqrt(pvar1)
    , tlo = newdat$W17Height-cmult*sqrt(tvar1)
    , thi = newdat$W17Height+cmult*sqrt(tvar1)
)

    
# With CI based on the fixed effect uncertainty and random effect variance
## CC families only
    F3 <- newdat %>%
      filter(Family == "CR10" | Family == "CR2" | Family =="CR3" | Family == "CR4" | Family == "CR5" | Family == "CR6" | Family == "CR7" | Family == "CR9") %>% 
    ggplot(aes(x=AHM, y=W17Height, color=Family))+
    geom_point()+
    stat_smooth(method="lm") +
    ylim(0,30) +
    geom_pointrange(aes(ymin = tlo, ymax = thi)) +
    scale_x_continuous(limits=c(16,30),breaks=c(17.5,23.44,27.54), labels=c("17.5", "23.4", "27.5")) +
    facet_wrap(~Family) +
    theme(legend.position="none") +
    labs(title="Figure 3: CI based on fixed effects and random effect variance")
    
    
    F4 <- newdat1 %>%
      filter(Family == "CR10" | Family == "CR2" | Family =="CR3" | Family == "CR4" | Family == "CR5" | Family == "CR6" | Family == "CR7" | Family == "CR9") %>% 
    ggplot(aes(x=PPT_wt, y=W17Height, color=Family))+
    geom_point()+
    stat_smooth(method="lm") +
    ylim(0,30) +
    geom_pointrange(aes(ymin = tlo, ymax = thi)) +
     scale_x_continuous(limits=c(175,350),breaks=c(182.4,218.9,342), labels=c("182.4", "218.9", "342")) +
    facet_wrap(~Family) +
    theme(legend.position="none") +
    labs(title="Figure 4: CI based on fixed effects and random effect variance")
```
  
# Results


Figure 1 and Figure 2 show the values for AHM and PPT_wt for the CR population families. CR4 has the smallest p-value for the interaction term in the regression model for AHM and PPT_wt (p = 0.09) (loess line shown not regression line). 

```{r, fig.height=4, fig.width=6}

F1

F2
```


Figure 3 shows the fitted regression lines for AHM for the CR population and Figure 4 shows the values for PPT_wt.

```{r, fig.height=4, fig.width=6}
F3

F4
```

Figure 5 compares the reaction norms of CR4 for AHM and PPT_wt. Note that these plots are flipped as the reaction norm depends just on the order of the sites as they align with the climate variables

```{r, fig.height=4, fig.width=8}

grid.arrange(F5a, F5b, ncol = 2)


```

Figure 6 and Figure 7 show the relationship between AHM and PPT_wt for Population with a loess line fitted. The BE population has the smallest p-value for the interaction term with AHM (p = 0.07) and PPT_wt (0.09).

```{r, fig.height=8, fig.width=6}

F6

F7
```

Figure 8 compares the reaction norms between AHM and PPT_wt for the BE Population - again note similarities due to planting site and climate variables being colinear.

``````{r, fig.height=4, fig.width=8}

grid.arrange(F8a, F8b, ncol = 2)
```

Table 1 shows the regression coefficients for AHM  and table 2 shows the coefficients  for PPT_wt.

```{r}


kable(summary(m3)$coef, caption="Table 1: Full results for regression model of Family and AHM on height", digits=2)

kable(summary(m3a)$coef, caption="Table 2: Full results for regression model of Family and PPT_wt on height", digits=2)

```

Table 3 shows the regression co-efficients for the effect of population on AHM and of population on PPT_wt in table 4.

```{r}

kable(summary(m4)$coef, caption="Table 3: Full results for regression model of Population and AHM on height", digits=2)

kable(summary(m5)$coef, caption="Table 4: Full results for regression model of Population and PPT_wt on height", digits=2)

```

